Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2019-03-06 mpaper/paper.tex
The notes PDF has been eliminated

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}

@@

\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}

@@

\newcommand{\dm}{\frac{\partial}{\partial\bm\mu}}
\newcommand{\dS}{\frac{\partial}{\partial\mathbf{S}}}
\newcommand{\dB}{\frac{\partial}{\partial\mathbf{B}}}
\newcommand{\dl}{\frac{\partial}{\partial \lambda_i}}
\newcommand{\dlj}{\frac{\partial}{\partial \lambda_j}}
\newcommand{\dt}{\frac{\partial}{\partial t}}
\newcommand{\df}{\left.\frac{\partial f}{\partial t}\right|_{(\mathbf{r},
    \mathbf{u}, t)}}

@@

\begin{abstract}
According to Simon Peyton Jones, an abstract should address
four key questions. First, what is the problem that this
paper tackles? Second, why is this an interesting problem?
Third, what is the solution this paper proposes?
Finally, why is the proposed solution a good one?
\end{abstract}

@@

As recently suggested by Ong et al. \cite{ong2018gaussian}, we use a
decomposition $\bm\Sigma = \mathbf{B}\mathbf{B}^\intercal + \mathbf{D}^2$, where
$\mathbf{B}$ is a lower triangular $m \times p$ matrix with positive diagonal
entries, and $\mathbf{D} = \diag(d_1, \dots, d_m)$. Typically, we would set $p$
so that $p \ll m$ to get an efficient approximation, but it is also worth
pointing out that we can retain full accuracy by setting $p = m$ and $\mathbf{D}
= \mathbf{O}$. Moreover, we define a few matrices that will simplify expressions
for the derivatives:
\begin{align*}
  \mathbf{U} &= (\mathbf{u} - \bm\mu)(\mathbf{u} - \bm\mu)^\intercal \\
  \mathbf{S} &= \Kru^\intercal\Kuu^{-1}, \\
  \bm\Gamma &= \Krr - \mathbf{S}\Kru, \\
  \mathbf{R} &= \mathbf{S}\frac{\partial \Kru}{\partial \lambda_i} - \frac{\partial \Krr}{\partial \lambda_i} + \left( \frac{\partial \Kru^\intercal}{\partial \lambda_i} - \mathbf{S}\frac{\partial \Kuu}{\partial \lambda_i} \right) \Kuu^{-1}\Kru.
\end{align*}

@@

\begin{lemma}[Derivatives of probability
  distributions] \label{lemma:derivatives}
  \leavevmode
  \begin{enumerate}
  \item $\frac{\partial q(\mathbf{u})}{\partial \bm\mu} =
    \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} + \bm\Sigma^{-\intercal})(\mathbf{u}
    - \bm\mu)$.
  \item
    \begin{enumerate}
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} =
      \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal})$.
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} =
      q(\mathbf{u})(\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
      |\bm\Sigma|^{-1}\adj(\bm\Sigma))\mathbf{B}$.
%    \item
%      $\frac{\partial q(\mathbf{u})}{\partial \mathbf{D}} =
%      \frac{1}{2}q(\mathbf{u}) \{ \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1}\mathbf{D}^\intercal +
%      \mathbf{D}^\intercal\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
%      |\bm\Sigma|^{-1}(\adj(\bm\Sigma^{-1})\mathbf{D}^\intercal +
%      \mathbf{D}^\intercal \adj(\bm\Sigma^{-1})) \}$.
    \end{enumerate}
  \item For $i = 0, \dots, d$,
    \[
      \begin{split}
        \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &=
        \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1}
          \tr(\mathbf{R} \adj(\bm\Gamma)) \\
          &- (\mathbf{r} -
          \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r}
          - \mathbf{Su})),
      \end{split}
    \]
  \end{enumerate}
\end{lemma}

@@

  If $i > 0$, then each element of $\frac{\partial \Kuu}{\partial
    \lambda_i}$ is a constant multiple of $k_{\bm\lambda}(\mathbf{x}_j,
  \mathbf{x}_k)$, for some $\mathbf{x}_j$ and $\mathbf{x}_k$. Since
  $k_{\bm\lambda}(\mathbf{x}_j, \mathbf{x}_k)$ is a decreasing function of
  $\lambda_i$, and $c(\mathbf{r}, \mathbf{u}) > a$,
  \[ %TODO: what the hell do I do here?
    \begin{split}
      k_{\bm\lambda}(\mathbf{x}_j, \mathbf{x}_k)|_{\lambda_i = 
        c(\mathbf{r}, \mathbf{u})} &=
      \begin{multlined}[t]
        \lambda_0 \exp \left( \vphantom{\sum_{n \in \{ \} \setminus}}
        \right. -\frac{1}{2}c(\mathbf{r}, \mathbf{u})(x_{j,i} - x_{k,i})^2 \\
        - \mathbbm{1}[j \ne k]\sigma^2c(\mathbf{r}, \mathbf{u}) \\
        - \left. \sum_{n \in \{ 1, \dots, d \} \setminus \{ i \}}
          \frac{1}{2} \lambda_n(x_{j,n} - x_{k,n})^2 + \mathbbm{1}[j \ne
          k]\sigma^2 \lambda_n \right)
      \end{multlined} \\
      &<
      \begin{multlined}[t]
        \lambda_0 \exp \left( \vphantom{\sum_{n \in \{ \} \setminus}}
        \right. -\frac{1}{2}a(x_{j,i} - x_{k,i})^2 -
        \mathbbm{1}[j \ne k]\sigma^2a \\
        - \left. \sum_{n \in \{ 1, \dots, d \} \setminus \{ i \}}
          \frac{1}{2} \lambda_n(x_{j,n} - x_{k,n})^2 + \mathbbm{1}[j \ne
          k]\sigma^2 \lambda_n \right),
      \end{multlined}
    \end{split}
  \]
  which gives an upper bound on each element of
  \[
    \left. \frac{\partial \Kuu}{\partial \lambda_i} \right|_{\lambda_i =
      c(\mathbf{r}, \mathbf{u})}.
  \]
  A similar line of reasoning establishes lower bounds as well.

@@

Our main theorem is a specialised version of an integral differentiation result
by Chen \cite{lecture_notes}.
\begin{theorem} \label{thm:main}
  Whenever the derivative exists,
  \[
    \dt\iint
    \V(s)q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})\dx
    = \iint
    \dt[\V(s)q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})]\dx,
  \]
  where $t$ is any scalar part of $\bm\mu$, $\bm\Sigma$, or $\bm\lambda$.
\end{theorem}
\begin{proof}
  Let
  \begin{align*}
    \f &= \V(s)q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u}), \\
    F(t) &= \iint \f\dx,
  \end{align*}
  and fix the value of $t$. Let $(t_n)_{n=1}^\infty$ be any sequence such that
  $\lim_{n \to \infty} t_n = t$, but $t_n \ne t$ for all $n$. We want to show
  that
  \begin{equation} \label{eq:to_prove}
    F'(t) = \lim_{n \to \infty} \frac{F(t_n) - F(t)}{t_n - t} = \iint \df\dx.
  \end{equation}
  We have
  \[
    \begin{split}
      \frac{F(t_n) - F(t)}{t_n - t} &= \iint \frac{\ftn - \f}{t_n - t}\dx \\
      &= \iint \fn\dx,
    \end{split}
  \]
  where
  \[
    \fn = \frac{\ftn - \f}{t_n - t}.
  \]
  Since
  \[
    \lim_{n \to \infty} \fn = \df,
  \]
  \eqref{eq:to_prove} follows from Theorem \ref{thm:lebesgue} as soon as we show
  that both $f$ and $f_n$ are measurable and find a non-negative integrable
  function $g$ such that for all $n$, $\mathbf{r}$, $\mathbf{u}$,
  \[
    |\fn| \le \g.
  \]
  The MDP value function is measurable by Proposition \ref{thm:measurability}.
  The result of multiplying or adding measurable functions (e.g., probability
  density functions (PDFs)) to a measurable function is still measurable. Thus,
  both $f$ and $f_n$ are measurable.

@@

>>>>

\documentclass{mpaper}
\usepackage{bm}
\usepackage{bbm}
\usepackage{mathtools}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}

@@

\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}
\newcommand{\Luu}{\mathbf{L}_{\mathbf{u},\mathbf{u}}}

@@

\newcommand{\dm}{\frac{\partial}{\partial\bm\mu}}
\newcommand{\dB}{\frac{\partial}{\partial\mathbf{B}}}
\newcommand{\dl}{\frac{\partial}{\partial \lambda_i}}
\newcommand{\dlj}{\frac{\partial}{\partial \lambda_j}}
\newcommand{\dt}{\frac{\partial}{\partial t}}
\newcommand{\df}{\left.\frac{\partial f}{\partial t}\right|_{(\mathbf{r},
    \mathbf{u}, t)}}

@@

\begin{abstract}
\end{abstract}

@@

As recently suggested by Ong et al. \cite{ong2018gaussian}, we use a
decomposition $\bm\Sigma = \mathbf{B}\mathbf{B}^\intercal + \mathbf{D}^2$, where
$\mathbf{B}$ is a lower triangular $m \times p$ matrix with positive diagonal
entries, and $\mathbf{D} = \diag(d_1, \dots, d_m)$. Typically, we would set $p$
so that $p \ll m$ to get an efficient approximation, but it is also worth
pointing out that we can retain full accuracy by setting $p = m$ and $\mathbf{D}
= \mathbf{O}$. Moreover, we define a few variables that will simplify
expressions for the derivatives:
\begin{align*}
  \mathbf{U} &= (\mathbf{u} - \bm\mu)(\mathbf{u} - \bm\mu)^\intercal \\
  \mathbf{S} &= \Kru^\intercal\Kuu^{-1}, \\
  \bm\Gamma &= \Krr - \mathbf{S}\Kru, \\
  \mathbf{R} &= \mathbf{S}\frac{\partial \Kru}{\partial \lambda_i} - \frac{\partial \Krr}{\partial \lambda_i} + \left( \frac{\partial \Kru^\intercal}{\partial \lambda_i} - \mathbf{S}\frac{\partial \Kuu}{\partial \lambda_i} \right) \Kuu^{-1}\Kru, \\
  Q &= (\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} - \bm\mu).
\end{align*}

@@

% TODO: define PDF before this point
\begin{lemma}[Derivatives of PDFs] \label{lemma:derivatives}
  \leavevmode
  \begin{enumerate}
  \item $\frac{\partial q(\mathbf{u})}{\partial \bm\mu} =
    \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} + \bm\Sigma^{-\intercal})(\mathbf{u}
    - \bm\mu)$.
  \item
    \begin{enumerate}
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} =
      \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal})$.
    \item
      $\frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} =
      q(\mathbf{u})(\bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} -
      |\bm\Sigma|^{-1}\adj(\bm\Sigma))\mathbf{B}$.
    \end{enumerate}
  \item For $i = 0, \dots, d$,
    \[
      \begin{split}
        \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} &=
        \frac{1}{2}q(\mathbf{r} \mid \mathbf{u}) (|\bm\Gamma|^{-1}
          \tr(\mathbf{R} \adj(\bm\Gamma)) \\
          &- (\mathbf{r} -
          \mathbf{Su})^\intercal\bm\Gamma^{-1}\mathbf{R}\bm\Gamma^{-1}(\mathbf{r}
          - \mathbf{Su})),
      \end{split}
    \]
  \end{enumerate}
\end{lemma}
\begin{proof}
  \leavevmode
  \begin{enumerate}
  \item
    \[
      \begin{split}
        \frac{\partial q(\mathbf{u})}{\partial m} &=
        q(\mathbf{u})\dm\left[-\frac{Q}{2}\right]
        \\
        &= -\frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu)\dm[\mathbf{u} -
        \bm\mu] \\
        &= \frac{1}{2}q(\mathbf{u})(\bm\Sigma^{-1} +
        \bm\Sigma^{-\intercal})(\mathbf{u} - \bm\mu).
      \end{split}
    \]
  \item
    \begin{align}
      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\exp \left( -\frac{Q}{2} \right)\right] \nonumber \\
                                                         &= \dB\left[\frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\right]\exp \left( -\frac{Q}{2} \right) \nonumber \\
                                                         &+ \frac{1}{(2\pi)^{m/2}|\bm\Sigma|^{1/2}}\dB\left[\exp\left( -\frac{Q}{2} \right)\right] \nonumber \\
                                                         &= -\frac{1}{2} q(\mathbf{u}) \left( |\bm\Sigma|^{-3/2} \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} + \frac{\partial Q}{\partial \mathbf{B}} \right). \label{eq:2_before}
    \end{align}
    For derivatives with respect to $\bm\Sigma$, we can refer to Petersen and
    Pedersen \cite{petersen2008matrix}:
    \begin{equation} \label{eq:partial_derivatives1}
      \frac{\partial |\bm\Sigma|}{\partial \bm\Sigma} = |\bm\Sigma|\bm\Sigma^{-\intercal}, \quad \frac{\partial Q}{\partial \mathbf{B}} = -\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal},
    \end{equation}
    while we can use an online tool by Laue et
    al.\footnote{\url{http://www.matrixcalculus.org/}}
    \cite{DBLP:conf/nips/LaueMG18} for the remaining ones:
    \begin{equation} \label{eq:partial_derivatives2}
      \begin{gathered}
        \frac{\partial |\bm\Sigma|}{\partial \mathbf{B}} =
        (\adj(\mathbf{T}) + \adj(\bm\Sigma))\mathbf{B}, \\
        \dB[(\mathbf{u} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{u} -
        \bm\mu)] = -(\mathbf{TUT} +
        \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1})\mathbf{B}.
      \end{gathered}
    \end{equation}
    Substituting results from \eqref{eq:partial_derivatives1} and
    \eqref{eq:partial_derivatives2} back into \eqref{eq:2_before} gives:
    \begin{alignat*}{3}
      \frac{\partial q(\mathbf{u})}{\partial \bm\Sigma} &=
      \frac{1}{2}q(\mathbf{u})(&&\bm\Sigma^{-\intercal}\mathbf{U}\bm\Sigma^{-\intercal}
      - \bm\Sigma^{-\intercal}), \\
      \frac{\partial q(\mathbf{u})}{\partial \mathbf{B}} &=
      \frac{1}{2}q(\mathbf{u}) \{&&|\bm\Sigma|^{-3/2}(\adj(\mathbf{T}) +
      \adj(\bm\Sigma)) \\
      & &&+ \mathbf{TUT} + \bm\Sigma^{-1}\mathbf{U}\bm\Sigma^{-1} \} \mathbf{B}.
    \end{alignat*}
  \item Using a result by Rasmussen and Williams % TODO: update the proof
                                % according to the new result
    \cite{DBLP:books/lib/RasmussenW06},
    \[
      \frac{\partial q(\mathbf{r} \mid \mathbf{u})}{\partial \lambda_i} =
      q(\mathbf{r} \mid \mathbf{u}) \dl
      \left[-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} -
        \frac{1}{2}\log|\Kuu| \right] = \frac{1}{2}q(\mathbf{r} \mid \mathbf{u})\tr
      \left((\Kuu^{-1}\mathbf{u}\mathbf{u}^\intercal\Kuu^{-\intercal} -
        \Kuu^{-1}) \frac{\partial \Kuu}{\partial \lambda_i} \right).
    \]
    The remaining derivative is
    \[
      \frac{\partial \Kuu}{\partial \lambda_i} =
      \begin{cases}
        \frac{1}{\lambda_i}\Kuu & \text{if } i = 0, \\
        \Luu & \text{otherwise,}
      \end{cases}
    \]
    where
    \[
      \begin{split}
        [\Luu]_{j,k} &= \dl k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j},
        \mathbf{x}_{\mathbf{u},k}) \\
        &= k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
        \dl \left[-\frac{1}{2}(\mathbf{x}_{\mathbf{u},j} -
          \mathbf{x}_{\mathbf{u},k})^\intercal \bm\Lambda
          (\mathbf{x}_{\mathbf{u},j} - \mathbf{x}_{\mathbf{u},k}) -
          \mathbbm{1}[j \ne k]\sigma^2\tr(\bm\Lambda) \right] \\
        &= k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
        \dl \left[-\frac{1}{2}\sum_{l=1}^d \lambda_l
          (x_{\mathbf{u},j,l} - x_{\mathbf{u},k,l})^2 -
          \mathbbm{1}[j \ne k]\sigma^2\sum_{l=1}^d \lambda_l \right] \\
        &= k_{\bm\lambda}(\mathbf{x}_{\mathbf{u},j}, \mathbf{x}_{\mathbf{u},k})
        \left( -\frac{1}{2}(x_{\mathbf{u},j,i} -
        x_{\mathbf{u},k,i})^2 - \mathbbm{1}[j \ne k]\sigma^2 \right).
      \end{split}
    \]
  \end{enumerate}
\end{proof}

@@

  If $i > 0$, then each element of $\frac{\partial \Kuu}{\partial
    \lambda_i}$ is a constant multiple of $k_{\bm\lambda}(\mathbf{x}_j,
  \mathbf{x}_k)$, for some $\mathbf{x}_j$ and $\mathbf{x}_k$. Since
  $k_{\bm\lambda}(\mathbf{x}_j, \mathbf{x}_k)$ is a decreasing function of
  $\lambda_i$, and $c(\mathbf{r}, \mathbf{u}) > a$,
  \begin{alignat*}{3}
    k_{\bm\lambda}(\mathbf{x}_j, \mathbf{x}_k)|_{\lambda_i = c(\mathbf{r},
      \mathbf{u})} &= \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. &&
    -\frac{1}{2} c(\mathbf{r}, \mathbf{u})(x_{j,i} - x_{k,i})^2 \\
    & && - \mathbbm{1}[j \ne k]\sigma^2c(\mathbf{r}, \mathbf{u}) - S \left.
      \vphantom{\frac{1}{2}} \right) \\
    &< \lambda_0 \exp \left( \vphantom{\frac{1}{2}} \right. && -\frac{1}{2}
    a(x_{j,i} - x_{k,i})^2 \\
    & && - \mathbbm{1}[j \ne k]\sigma^2a - S \left. \vphantom{\frac{1}{2}}
    \right),
  \end{alignat*}
  where
  \[
    S = \sum_{n \in \{ 1, \dots, d \} \setminus \{ i \}} \frac{1}{2}
    \lambda_n(x_{j,n} - x_{k,n})^2 + \mathbbm{1}[j \ne k]\sigma^2 \lambda_n,
  \]
  which gives an upper bound on each element of
  \[
    \left. \frac{\partial \Kuu}{\partial \lambda_i} \right|_{\lambda_i =
      c(\mathbf{r}, \mathbf{u})}.
  \]
  A similar line of reasoning establishes lower bounds as well.

@@

Our main theorem is a specialised version of an integral differentiation result
by Chen \cite{lecture_notes}.
\begin{theorem} \label{thm:main}
  Whenever the derivative exists,
  \[
    \dt\iint
    \V(s)q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})\dx
    = \iint
    \dt[\V(s)q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u})]\dx,
  \]
  where $t$ is any scalar part of $\bm\mu$, $\bm\Sigma$, or $\bm\lambda$.
\end{theorem}
\begin{proof}
  Let
  \begin{align*}
    \f &= \V(s)q(\mathbf{r} \mid \mathbf{u})q(\mathbf{u}), \\
    F(t) &= \iint \f\dx,
  \end{align*}
  and fix the value of $t$. Let $(t_n)_{n=1}^\infty$ be any sequence such that
  $\lim_{n \to \infty} t_n = t$, but $t_n \ne t$ for all $n$. We want to show
  that
  \begin{equation} \label{eq:to_prove}
    F'(t) = \lim_{n \to \infty} \frac{F(t_n) - F(t)}{t_n - t} = \iint \df\dx.
  \end{equation}
  We have
  \[
    \begin{split}
      \frac{F(t_n) - F(t)}{t_n - t} &= \iint \frac{\ftn - \f}{t_n - t}\dx \\
      &= \iint \fn\dx,
    \end{split}
  \]
  where
  \[
    \fn = \frac{\ftn - \f}{t_n - t}.
  \]
  Since
  \[
    \lim_{n \to \infty} \fn = \df,
  \]
  \eqref{eq:to_prove} follows from Theorem \ref{thm:lebesgue} as soon as we show
  that both $f$ and $f_n$ are measurable and find a non-negative integrable
  function $g$ such that for all $n$, $\mathbf{r}$, $\mathbf{u}$,
  \[
    |\fn| \le \g.
  \]
  The MDP value function is measurable by Proposition \ref{thm:measurability}.
  The result of multiplying or adding measurable functions (e.g., probability
  density functions) to a measurable function is still measurable. Thus,
  both $f$ and $f_n$ are measurable.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{elbo_over_mu}
\end{figure*}
